---
title: "Project 1"
author: "Keerthana Radhakrishnan"
date: "1/15/2020"
output:
  word_document: default
  html_document: default
---
```{r}
sat<-read.csv("SATdata.csv",header=T,sep=",",as.is=T)
```
Exploratory data analysis: scatterplots
```{r}
par(mfrow=c(1,2))
plot(SATscore~Salary,data=sat,main="SATscore vs salary")
lines(lowess(sat$Salary,sat$SATscore),col="green")
plot(SATscore~Percent,data=sat,main="SATscore vs percent")
lines(lowess(sat$Percent,sat$SATscore),col="red")
```
Interpretation:
Both salary and percent do not have a linear association with the SATscore i.e.,both scatterplots suggests to include quadratic terms for salary and percent in the regression model. 

(b) Fitting the model
```{r}
mod<-lm(SATscore~Salary+Percent,data=sat)
summary(mod)
```
Checking the model assumptions
```{r}
par(mfrow=c(2,2))

qqnorm(rstandard(mod),main="qqplot",cex.main=0.8)
qqline(rstandard(mod))
plot(rstandard(mod)~fitted(mod),col=124,main="standardized residual&fitted",cex.main=0.8)
lines(lowess(rstandard(mod)~fitted(mod)))
plot(rstandard(mod)~sat$Salary,main="standardized residuals&salary",cex.main=0.8)
lines(lowess(rstandard(mod)~sat$Salary))
plot(rstandard(mod)~sat$Percent,main="standardized residuals&percent",cex.main=0.8)
lines(lowess(rstandard(mod)~sat$Percent))
```
Interpretation:
The qq plot shows that the data is normally distributed.
The standardised residuals vs fitted model shows non-linearity with a possible right-skewed data. 
Both standardised residuals vs salary and standardised residuals vs percent show non-linear trend in the plots.
```{r}
library(nortest)
ad.test(rstandard(mod))
```
As the p value in the above test (0.60) is greater than the significance level 0.05, we can say that there is an evidence that the data is normal.

(c) quadratic equation model
```{r}
Percent2<-sat$Percent^2
mod.quad<-lm(SATscore~Salary+Percent+Percent2,data=sat)
summary(mod.quad)
```
```{r}
par(mfrow=c(2,3))
qqnorm(rstandard(mod.quad),main="qqplot for mod.quad",cex.main=0.8)
qqline(rstandard(mod.quad))
plot(rstandard(mod.quad)~fitted(mod.quad),col=124,main="standardized residual&fitted",cex.main=0.8)
lines(lowess(rstandard(mod.quad)~fitted(mod.quad)))
plot(rstandard(mod.quad)~sat$Salary,main="standardized residuals&salary",cex.main=0.8)
lines(lowess(rstandard(mod.quad)~sat$Salary))
plot(rstandard(mod.quad)~sat$Percent,main="standardized residuals&percent",cex.main=0.8)
lines(lowess(rstandard(mod.quad)~sat$Percent))
plot(rstandard(mod.quad)~Percent2,main="standardized residuals&percent2",cex.main=0.8)
lines(lowess(rstandard(mod.quad)~Percent2))

library(nortest)
ad.test(rstandard(mod.quad))
```
The Anderson-Darling Test is still greater than 0.05, which ensures the normality of the model. The qqplot also shows that the data is normally distributed.
The residuals vs fitted plot is slightly deviated from the straight line, hence, showing slight non-linearity. 
While plots for standard residuals and percent related terms show linearity, the plot of standard residuals against salary still shows non- linearity.

(d) including quadratic terms for both Salary and Percent variables
```{r}
Salary2<-sat$Salary^2
mod.quad2<-lm(SATscore~Salary+Percent+Percent2+Salary2,data=sat)
summary(mod.quad2)
```
From the above R output, we can see that the p value for 'Salary2', the quadratic term of Salary is 0.22, whch is greater than the level of significance 0.05. Hence, we can say that this quadratic term does not have a significant effect on the SATscore. Hence, we can exclude Salary2 from the model.

(e) Interaction between salary and percent
```{r}
mod.int<-lm(SATscore~Salary+Percent+Salary*Percent+Salary*Percent2,data=sat)
summary(mod.int)
anova(mod.quad,mod.int)
```
From the above R output, we can see that both the interaction terms i.e., Salary:Percent and Salary:Percent2 have p values greater than the level of significance 0.05. Also, the F value is 0.195, which is also greater than 0.05. Hence, we can conclude that there is no statistically significant difference between the two models, mod.quad and mod.int.  

f) The final model will be:
mean(SATscore) = beta0 + beta1(Salary) + beta2(Percent) + beta3(Percent2)
```{r}
summary(mod.quad)
confint(mod.quad)
```
            point Estimates   Conf.int        Std.error     p-value       
Salary       1.803e+00        (0.19,3.4)      7.998e-01     0.029 *  
Percent     -6.646e+00        (-8.06,-5.2)    7.028e-01     2.33e-12 ***
Percent2     5.075e-02        (0.03,0.06)     8.924e-03     8.51e-07 ***


(g)
```{r}
library(gmodels)
estimable(mod.quad,c(1,25,5,25),conf.int = 0.95)
estimable(mod.quad,c(1,25,30,900),conf.int = 0.95)
estimable(mod.quad,c(1,25,70,4900),conf.int = 0.95)

estimable(mod.quad,c(1,35,5,25),conf.int = 0.95)
estimable(mod.quad,c(1,35,30,900),conf.int = 0.95)
estimable(mod.quad,c(1,35,70,4900),conf.int = 0.95)

estimable(mod.quad,c(1,45,5,25),conf.int = 0.95)
estimable(mod.quad,c(1,45,30,900),conf.int = 0.95)
estimable(mod.quad,c(1,45,70,4900),conf.int = 0.95)
```
     Salary           Percent
                   5%    30%    70%
    25,000   |    1052   930    867
    35,000   |    1070   948    885
    45,000   |    1088   966    903

