---
title: "Project 4"
author: "Keerthana Radhakrishnan"
date: "3/2/2020"
output: word_document
---
```{r}
require(Hmisc)
require(multcomp)
require(nlme)
require(mosaic)
```



```{r}
osteo<-read.csv("Osteocalcindata.csv",as.is=TRUE,sep=",",header=TRUE)
```

```{r}
attach(osteo)
ost1<-reshape(osteo,varying=c("OC0","OC1","OC2","OC3"),v.names="OC",times = c(0,0.5,1,1.5),timevar = "dose",idvar = "id",new.row.names = 1:length(c(OC0,OC1,OC2,OC3)),direction="long",drop="ID1")
ost<-ost1[order(ost1$id),]
ost<-ost[,c(5,1,2,3,4)]
rownames(ost) <- NULL
ost
```
```{r}
male<-ost[which(ost$Sex=="M"),]
female<-ost[which(ost$Sex=="F"),]
```



**Exploratory data analysis:**
1. summary statistics:
```{r}
attach(ost)
t1 <- favstats(OC~Sex)[,-c(3,5,10)]
t2 <- favstats(OC0~Sex, data = osteo)[,-c(3,5,10)]
t3 <- favstats(OC1~Sex, data = osteo)[,-c(3,5,10)]
t4 <- favstats(OC2~Sex, data = osteo)[,-c(3,5,10)]
t5 <- favstats(OC3~Sex, data = osteo)[,-c(3,5,10)]
tt <- rbind(t1,t2,t3,t4,t5)

rownames(tt) <- c("Total-Female","Total-Male","OC0-Female","OC0-Male","OC1-Female","OC1-Male",
                  "OC2-Female","OC2-Male","OC3-Female","OC3-Male")

tt
```
2. Profile plot:
```{r}
#Profile Plot of Males
plot1 <- ggplot(male, aes(dose, OC, group=id))+
  geom_line(col="blue")+
  ylim(0,40)+
  ggtitle("Spaghetti plot for Males")

#Profile Plot of Females
plot2 <- ggplot(female, aes(dose, OC, group=id))+
  geom_line(col="red")+
  ylim(0,40)+
  ggtitle("Spaghetti plot for Females")

##Profile Plot of Data with Mean lines
plot3 <-  ggplot(ost, aes(dose, OC, group=id))+
  geom_line(aes(col=Sex))+
  stat_smooth(aes(group=Sex, col=Sex),se=F,size=2)+
  stat_summary(aes(group=Sex,lty=Sex),geom="point",fun.y = mean,size=2)+
  theme_bw()+
  ylim(0,40)+
  ggtitle("Sphagetti plot")
plot1
plot2
plot3

```
Based on the above spaghetti plot, we can expect differences in the mean osteocalcin levels for males and females at doses 0 and 0.5, but there might be no no/slight difference in the mean difference of males and females at doses 1.0 and 1.5.

**(b) Model building:**
```{r}
attach(ost)
mod1 <- lme(OC~Male * dose, random = ~dose|id, method = "ML")
summary(mod1)
# Independent model
mod.ind <- gls(OC~Male * dose)
m2 <- summary(mod.ind)$AIC

#Compound Symmetry
mod.cs <- gls(OC ~ Male * dose, correlation = corCompSymm(form = ~1|id))
m3 <- summary(mod.cs)$AIC

# AR1 
mod.ar1 <- gls(OC ~ Male * dose, correlation = corAR1(form = ~1|id))
m4 <- summary(mod.ar1)$AIC

# MA 1
mod.ma1 <- gls(OC ~ Male *dose, correlation = corARMA(p=0,q=1,form = ~1|id))
m5 <- summary(mod.ma1)$AIC


# ARMA 11
mod.arma <- gls(OC~ Male * dose, correlation = corARMA(p=1,q=1,form = ~1|id))
m6 <- summary(mod.arma)$AIC

# Unstructured
mod.unstr <- gls(OC ~ Male * dose, correlation = corSymm(form = ~1|id))
m7 <- summary(mod.unstr)$AIC


# Selected model uses the unstructured correlation matrix and considers different variance across group.

corr_mat<- rbind(m2,m3,m4,m5,m6,m7)
rownames(corr_mat) <- c("Independent","Compound Symmetry","AR1","MA1","ARMA (1,1)", "Unstructured")
colnames(corr_mat) <- "AIC"
corr_mat
```
The AIC is least for the unstructured correlation matrix which is 470.12. 
```{r}
# Unstructured correlation matrix: Different variance across dose
mod.var.dose <- gls(OC~ Sex * dose, correlation = corSymm(form = ~1|id), weights = varIdent(form = ~1|dose))
m8 <- summary(mod.var.dose)$AIC

# Unstructured correlation matrix: Different variance across Groups 
mod.var.group <- gls(OC ~ Sex * dose, correlation = corSymm(form = ~1|id), weights = varIdent(form = ~1|Sex))
m9 <- summary(mod.var.group)$AIC

unstr_var<-rbind(m8,m9)
colnames(unstr_var)<-"AIC"
rownames(unstr_var)<-c("Unstructured: Different variance across Dose","Unstructured: Different Variance across Group")
unstr_var
```
The preferred correlation matrix that will be used by the model is the unstructured correlation matrix with different variance across groups.

**(c) Modelling the Mean:**

*(A) Linear Trends with Different Intercepts but same slopes*
```{r}

mod.same <-  gls(OC~ Male + factor(dose), method= "ML",correlation = corSymm(form = ~1|id), weights = varIdent(form = ~1|Sex))
summary(mod.same)
```
*(B) Linear Trends with Different Intercepts and Different Slopes*
```{r}
mod.diff <- gls(OC ~ Male * factor(dose), method= "ML",correlation = corSymm(form = ~1|id), weights = varIdent(form = ~1|Sex))
summary(mod.diff)
```
*(C) Seperate Quadratic Trends for each Group*
```{r}
ost$sq.dose <- ost$dose^2
ost$dose <- factor(ost$dose)
ost$sq.dose <- factor(ost$sq.dose)
mod.quad<- gls(OC ~ Male*sq.dose, data=ost, method= "ML",correlation = corSymm(form = ~1|id), weights = varIdent(form = ~1|Sex))
summary(mod.quad)
```
**(d)The final model will be:**
$$y_{ij} = \beta_0 + \beta_1\text{Male}_i + \beta_2D_2 + \beta_3D_3 + \beta_4D_4 + \beta_5\text{Male*D2} + \beta_6\text{Male*D3} + \beta_7\text{Male*D4} + \epsilon_{ij}$$
$$ y_{ij} = \begin{bmatrix} y_{i1} \\ y_{i2} \\ y_{i3} \\ y_{i4} \end{bmatrix} \overset{\text{ind}} \sim MVN \left( \begin{bmatrix} \mu_{i1} \\ \mu_{i2} \\ \mu_{i3} \\ \mu_{i4} \end{bmatrix}, \Sigma_{4\times4} \right)   $$
$$\Sigma = \begin{pmatrix}  \sigma_{11} & \sigma_{12} & \sigma_{13} & \sigma_{14} \\ & \sigma_{22} & \sigma_{23} & \sigma_{24} \\ &  & \sigma_{33} & \sigma_{34} \\ & & &  \sigma_{44} \end{pmatrix}$$
$$\mu_{i1} = \beta_0 + \beta_1\text{Male}_i$$
$$\mu_{i2} = \beta_0 + \beta_1\text{Male}_i + \beta_2D_2 + \beta_5\text{Male}\times D2 $$
$$\mu_{i3} =  \beta_0 + \beta_1\text{Male}_i  + \beta_3D_3 +  \beta_6\text{Male} \times D3$$
$$\mu_{i4} = \beta_0 + \beta_1\text{Male}_i  + \beta_4D_4  + \beta_7\text{Male} \times D4$$
```{r}
summary(mod.diff)
c<-intervals(mod.diff)
c
```
*(e)Testing for slope of Females means:*
$$H_o : \beta_2 + \beta_3 + \beta_4 = 0$$

```{r}
mod.fem <- glht(mod.diff,matrix(c(0,0,1,1,1,0,0,0),1, length(coef(mod.diff))) )
summary(mod.fem)
confint(mod.fem)
```
*testing significance of Slope of males:*
$$H_o : \beta_0 + \beta_1 + \beta_2 + \beta_3 + \beta_4 + \beta_5 + \beta_6+ \beta_7= 0$$

```{r}
mod.m<- glht(mod.diff,linfct = matrix(c(0,1,1,1,1,1,1,1),1, length(coef(mod.diff))) )
summary(mod.m)
confint(mod.m)
```
(f)
```{r}
difference<-  glht(mod.diff,matrix(c(0,0,0,0,-1,0,0,-1),1, length(coef(mod.diff))))
confint(difference)
```
*(g)*
```{r}

male0<- glht(mod.diff,matrix(c(1,1,0,0,0,0,0,0),1, length(coef(mod.diff))))
male05<-glht(mod.diff,matrix(c(1,1,1,0,0,1,0,0),1, length(coef(mod.diff))))
male10<- glht(mod.diff,matrix(c(1,1,0,1,0,0,1,0),1, length(coef(mod.diff))))
male15<- glht(mod.diff,matrix(c(1,1,0,0,1,0,0,1),1, length(coef(mod.diff))))
fm0<- glht(mod.diff,matrix(c(1,0,0,0,0,0,0,0),1, length(coef(mod.diff))))
fm05<- glht(mod.diff,matrix(c(1,0,1,0,0,0,0,0),1, length(coef(mod.diff))))
fm10<- glht(mod.diff,matrix(c(1,0,0,1,0,0,0,0),1, length(coef(mod.diff))))
fm15<- glht(mod.diff,matrix(c(1,0,0,0,1,0,0,0),1, length(coef(mod.diff))))


dif0<- glht(mod.diff,matrix(c(0,1,0,0,0,0,0,0),1, length(coef(mod.diff))))
dif05<- glht(mod.diff,matrix(c(0,1,0,0,0,1,0,0),1, length(coef(mod.diff))))
dif10<- glht(mod.diff, matrix(c(0,1,0,0,0,0,1,0),1, length(coef(mod.diff))))
dif15<- glht(mod.diff,matrix(c(0,1,0,0,0,0,0,1),1, length(coef(mod.diff))))

x1 <- confint(male0)$confint
x2 <- confint(male05)$confint
x3 <- confint(male10)$confint
x4 <- confint(male15)$confint
x5 <- confint(fm0)$confint
x6 <- confint(fm05)$confint
x7 <- confint(fm10)$confint
x8 <- confint(fm15)$confint
x9 <- confint(dif0)$confint
x10 <- confint(dif05)$confint
x11 <- confint(dif10)$confint
x12 <- confint(dif15)$confint

est <- rbind(x1[1,],x2[1,],x3[1,],x4[1,],x5[1,],x6[1,],x7[1,],x8[1,],x9[1,],x10[1,],x11[1,],x2[1,])


rownames(est) <- c("Male at 0","Male at 0.5","Male at 1","Male at 1.5","Female at 0","Female at 0.5","Female at 1","Female at 1.5","(Male-Female) at 0",
                  "(Male-Female) at 0.5","(Male-Female) at 1.0","(Male-Female) at 1.5")

est
```
(h)

